\documentclass[10pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}

\title{Vision-Only Latent Diffusion Super-Resolution without T2I Pretraining}
\author{sr-diffusion contributors}
\date{Completed-experiment snapshot through Stage 2 XL candidate selection}

\begin{document}
\maketitle

\begin{abstract}
This report summarizes a work-in-progress x4 super-resolution system trained
directly in vision space, without importing a pretrained text-to-image diffusion
backbone. The completed pipeline consists of a factor-4 VAE, deterministic
LR-to-HR-latent pretraining, conditional latent diffusion, and
condition-start fine-tuning on a 103k-image photo split. The latest completed
XL-stage result is a 500M-class condition encoder trained with a stronger
degradation curriculum. Its Stage 2 candidates were compared on the same fixed
validation images, and the 72k-step checkpoint was selected for the next
diffusion stage.
\end{abstract}

\section{Objective}

The target task is single-image x4 restoration:
\[
128 \times 128 \rightarrow 512 \times 512.
\]
A later extension targets
\[
192 \times 192 \rightarrow 768 \times 768.
\]
The design goal is a practical latent diffusion SR model trained from project
checkpoints only. The system intentionally avoids Stable Diffusion, T2I
backbones, text conditioning, and imported generative priors from pretrained
diffusion models.

\section{Architecture}

The model is split into three trainable parts:
\[
\mathrm{HR\ image} \rightarrow \mathrm{VAE} \rightarrow \mathrm{HR\ latent},
\]
\[
\mathrm{LR\ image} \rightarrow \mathrm{condition\ encoder} \rightarrow
\mathrm{condition\ latent},
\]
\[
\mathrm{noisy\ HR\ latent}, \mathrm{condition\ latent}, t,
\mathrm{domain\ id} \rightarrow \mathrm{conditional\ U\mbox{-}Net}.
\]
The VAE decoder maps the denoised latent back to the 512px HR output. Domain IDs
are implemented for photo and anime/illustration data, although the completed
large-scale runs reported here use the photo split.

\section{Training Stages}

The project is organized as staged training rather than a single end-to-end run:

\begin{enumerate}
  \item Stage 1 trains the factor-4 VAE on HR crops.
  \item Stage 2 trains a deterministic condition encoder to predict HR latents
        from degraded LR inputs.
  \item Stage 3 trains a conditional diffusion U-Net in latent space.
  \item Stage 4 fine-tunes diffusion from the condition latent at low timesteps.
  \item XL variants widen and deepen the condition encoder or diffusion U-Net
        while using partial checkpoint initialization where tensor shapes match.
\end{enumerate}

\section{Completed Implementation}

The repository currently includes:
\begin{itemize}
  \item manifest-based dataset loading with explicit domain IDs,
  \item on-the-fly x4 degradation pipelines,
  \item a factor-4 AutoencoderKL with 16 latent channels,
  \item deterministic LR-to-HR-latent pretraining,
  \item conditional latent diffusion with timestep and domain conditioning,
  \item DDIM/img2img sampling for validation,
  \item condition-start Stage 4 fine-tuning,
  \item bf16 CUDA/ROCm training paths,
  \item W\&B logging, fixed validation samples, checkpoint evaluation, and
        recovery scripts for VM migration,
  \item partial checkpoint initialization for widened/deepened models.
\end{itemize}

\section{Data and Degradation}

The large-scale photo runs use a 103,450-image training split and a fixed
100-image validation split. HR crops are 512px and LR inputs are 128px. The
original photo degradation was extended with stronger blur, noise, compression,
ringing, color shift, and banding variants. The latest XL condition encoder uses
the \texttt{photo\_v3\_noise\_mix} degradation preset to emphasize
denoise/sharpening behavior.

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
Item & Value \\
\midrule
Task & x4 SR, 128px LR to 512px HR \\
Active large-scale domain & Photo \\
Train split & 103,450 images \\
Validation split & 100 fixed images \\
VAE latent & 128px, 16 channels \\
Primary dtype & bf16 \\
Latest completed XL degradation & \texttt{photo\_v3\_noise\_mix} \\
\bottomrule
\end{tabular}
\end{table}

\section{Completed Results}

\subsection{Stage 1 VAE}

The selected Stage 1 VAE checkpoint is
\texttt{autoencoder\_photo10k\_b16\_eval\_online/checkpoints/best\_eval\_recon.pt}.
It was selected at the 50k-step checkpoint with:

\begin{table}[h]
\centering
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
eval/recon & 0.01198 \\
eval/kl & 9.38684 \\
eval/psnr & 40.19 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Photo100k Baseline Stages}

The first photo100k scale-up completed deterministic latent pretraining,
conditional diffusion, and condition-start fine-tuning.

\begin{table}[h]
\centering
\begin{tabular}{llll}
\toprule
Stage & Run & Finished step & Selected result \\
\midrule
Stage 2 & \texttt{latent\_pretrain\_photo100k\_b64} & 30k &
best latent loss 0.21230 \\
Stage 3 & \texttt{diffusion\_photo100k\_b32} & 60k &
decoded PSNR 24.56 diagnostic \\
Stage 4 & \texttt{diffusion\_photo100k\_b32\_stage4\_condition} & 5k &
sampled val100 PSNR 25.4072 \\
\bottomrule
\end{tabular}
\end{table}

On the sampled validation comparison, the Stage 4 condition-start checkpoint
slightly improved the Stage 3 checkpoint:
\[
25.4072 - 25.3745 = +0.0327\ \mathrm{PSNR}.
\]

\subsection{Photo100k v2 Denoise/Sharpening Stages}

The \texttt{photo\_v2} degradation curriculum produced lower absolute PSNR than
the earlier degradation because the task is stronger, but Stage 4 improved both
the Stage 3 v2 model and bicubic baseline on the sampled validation set.

\begin{table}[h]
\centering
\begin{tabular}{lr}
\toprule
Comparison & Sampled val100 PSNR \\
\midrule
Bicubic baseline & 22.4103 \\
Stage 3 v2 SR & 22.6699 \\
Stage 4 v2 SR & 22.8426 \\
\bottomrule
\end{tabular}
\end{table}

Stage 4 v2 improved over Stage 3 v2 by +0.1727 PSNR and beat bicubic by +0.4323
PSNR on the sampled validation set, with 81 wins and 19 losses in per-image
comparison. Remaining artifacts included occasional color/contrast overshoot and
small cyan/green sampling artifacts.

\subsection{Stage 2 XL Condition Encoder}

A smaller \texttt{photo\_v3\_noise\_mix} Stage 2 run was stopped at step 12,700
after the validation latent loss plateaued near 0.282. The XL condition encoder
then completed 80k steps with a larger 500M-class full-path target. Its main
checkpoints were:

\begin{table}[h]
\centering
\begin{tabular}{llll}
\toprule
Checkpoint & Step & eval/latent\_loss & eval/decoded\_psnr \\
\midrule
Best latent & 66k & 0.27230 & 21.38 \\
Best PSNR proxy & 72k & 0.27940 & 21.52 \\
Final/latest & 80k & 0.27592 & 21.51 \\
\bottomrule
\end{tabular}
\end{table}

The three Stage 2 XL candidates were then compared on the same 100 validation
images by decoding condition-only outputs through the selected VAE. This
comparison used fixed samples, fixed seed, and the same
\texttt{latent\_pretrain\_photo100k\_v3\_noise\_xl.yaml} configuration.

\begin{table}[h]
\centering
\begin{tabular}{lrr}
\toprule
Candidate & Mean decoded MSE & Mean decoded PSNR \\
\midrule
Best latent, step 66k & 0.029092 & 22.8017 \\
Latest, step 80k & 0.028277 & 22.8901 \\
Best PSNR proxy, step 72k & 0.028161 & 22.9029 \\
Bicubic reference & -- & 22.3599 \\
\bottomrule
\end{tabular}
\end{table}

The 72k-step checkpoint had the best decoded condition-only PSNR and the lowest
decoded MSE, so it was selected as the condition encoder for the next XL
diffusion experiment.

\section{Current Scope of This Report}

This document records completed experiments only. XL diffusion training and any
metrics from runs still in progress are intentionally excluded until those runs
finish and their checkpoints can be compared on fixed validation samples.

\section{Limitations}

This is an engineering report for an active experiment, not a claim of a final
published method. The large-scale completed results are photo-domain only, while
anime/illustration domain conditioning remains implemented but not yet validated
at the same scale. The reported sampled PSNR values are useful for internal
checkpoint comparison, but they are not a full benchmark suite against external
SR methods.

\begin{thebibliography}{9}
\bibitem{rombach2022latent}
Robin Rombach, Andreas Blattmann, Dominik Lorenz, Patrick Esser, and Bjorn
Ommer. High-Resolution Image Synthesis with Latent Diffusion Models. CVPR,
2022.

\bibitem{saharia2022sr3}
Chitwan Saharia et al. Image Super-Resolution via Iterative Refinement. IEEE
TPAMI, 2022.
\end{thebibliography}

\end{document}
